Modular flow and holographic entanglement entropy in cutoff AdS

### 2021-05-31 alpha/CutoffModularFlow/cutoffModular-alpha.tex
Update note on cutoff modular flow
- add comments
- tweak notations

	\item We then note that the periodicity can be restored by following the flow of $\xi$ starting from a codim-2 cutoff surface. This could be related to the \textit{restricted maximin} proposal of \cite{Marolf:2019bgj,Grado-White:2020wlb}.

@@

\section{Modular flow in the Poincar\'e patch}

\subsection{Review of the general recipe}
	A boundary interval $\mcal{A}$ defines a denisty matrix $\rho_\mcal{A} = \tr_{\bar{\mcal{A}}} \rho = e^{-\mcal{H}_\mcal{A}}$. $\mcal{H}_\mcal{A}$ is generically nonlocal, but in a highly symmetric background (e.g.~AdS vacuum) it might be geometrically realized and corresponds to a symmetry of the underlying quantum field theory. This is the modular flow $\xi$.

	In such case there is a generalized Rindler transformation $f$ that maps the casual domain $\mcal{D}$ to a patch $\{(\tau,\tilde{x})\}$ in some noncompact generalized Rindler spacetime, with $\pdd{\tau}$ an isometry of the spacetime. Then we have:
	\begin{equation}
		\xi = 2\pi \pdd{\tau}
		= 2\pi \sum_i a^i H_i,
	\quad
		\tau \sim \tau + 2\pi i
	\end{equation}
	Here $H_i$ are the symmetry generators. The causal domain of dependence $\mcal{D}$ is mapped to some thermal state with normalized inverse temperature $2\pi$. Here we choose to absorb such $2\pi$ into the definition of $\xi$, so the Euclidean periodicity w.r.t.~$\xi$ is simply $\beta_\xi = 1 = \frac{1}{T_\xi}$, which means the surface gravity along the horizon is given by $\kappa_\xi = 2\pi T_\xi = 2\pi$.

@@

\subsubsection{Matching bulk and boundary symmetries}
	In Poincar\'e $\mrm{AdS}_3/\mrm{CFT}_2$, the $H_i$'s are given by $L_n,\bar{L}_n$ with $n=0,\pm 1$. At $z\to 0$ we recover the usual global $\mfrak{sl}(2,\mbb{R})$ generators:
	\begin{equation}
		z\to 0,
	\quad
		      L_n \to -u^{n+1} \pdd{u},\quad
		\bar{L}_n \to -v^{n+1} \pdd{v},\quad
	n = 0,\pm 1
	\end{equation}

	To extend it to the bulk, think of the corresponding bulk isometries: translations, rotations, dilations and special conformal transformations:
	\begin{equation}
	\begin{aligned}
		P_\mu
		&= \pdd{\mu},\\
		m_{uv}
		&= x_\mu \pdd{\nu} - x_\nu \pdd{\mu}
		= u\pdd{u} - v\pdd{v},\\
		\Delta
		&= X^I \pdd{I}
		= u\pdd{u} + v\pdd{v} + z\pdd{z},\\
		K_\mu
		&= (z^2 + uv)\,\pdd{\mu}
			- 2\eta_{\mu\nu} x^\nu \Delta,
	\end{aligned}
	\end{equation}
	\begin{equation}
		K_u = (z^2 + uv)\,\pdd{u} - v \Delta,
	\quad
		K_v = (z^2 + uv)\,\pdd{v} - u \Delta
	\end{equation}
	By requiring that a linear combination of these generators reduce to the boundary $L_n$'s, one can uniquely extend the $L_n,\bar{L}_n$'s into the bulk. The result for $L_n$'s is listed below; for $\bar{L}_n$ one need only exchange all $u\leftrightarrow v$.
	\begin{equation}
		L_{-1} = -\pdd{u},
	\quad
		L_0 = -u\,\pdd{u} - \frac{z}{2}\,\pdd{z},
	\quad
		L_1 = -u^2\pdd{u} + z^2 \pdd{v}
			- uz\,\pdd{z}
	\end{equation}

@@

	The RT surface (spacelike geodesic of Poincar\'e AdS with both ends attached to the boundary) is parametrized in terms of its end points $(t,x)_\pm$; it's the intersection of a AdS ``sphere'' (hyperboloid) and a plane through its center:
	\begin{equation}
	\text{``sphere'':}\quad\mathsmaller{
		z^2
		+ \pqty{x - \frac{x_+ + x_-}{2}}^{\!\!2}
		- \pqty{t - \frac{t_+ + t_-}{2}}^{\!\!2}
		= \pqty{\frac{L_\infty}{2}}^{\!\!2}
		= \pqty{\frac{x_+ - x_-}{2}}^{\!\!2}
		- \pqty{\frac{t_+ - t_-}{2}}^{\!\!2}
	}
	\end{equation}
	\begin{equation}
		L^2
		= \pqty{x_+ - x_-}^2
			- \pqty{t_+ - t_-}^2
		= (u_+ - u_-)(v_+ - v_-)
	\end{equation}
	\begin{equation}
	\text{plane:}\quad
		\frac{x - \frac{1}{2} (x_+ + x_-)}{x_+ - x_-}
		= \frac{t - \frac{1}{2} (t_+ + t_-)}{t_+ - t_-}
	\end{equation}
	Alternatively, it can be more compactly rewritten with $x^\mu \sim (t,x)$ or $(u,v)$; mind the various $\pm$ signs in the above expressions: 
	\begin{equation}
	\begin{aligned}
		0 &= z^2
			+ \eta_{\mu\nu} (x - x_+)^\mu (x - x_-)^\nu \\
		& = z^2
			+ (x - x_+)(x - x_-)
			- (t - t_+)(t - t_-) \\
		&= z^2 + \frac{
				(u - u_+)(v - v_-)
				+ (v - v_+)(u - u_-)
			}{2}
	\end{aligned}
	\end{equation}

	The bulk modular generator vanishes along the RT surface, i.e.~the RT surface is the fixed point of the modular flow; with this condition we can fix the coefficients of the modular generator, up to an overall coefficient\footnote{
		One can compare this with similar results in the literature such as \cite{Lashkari:2016idm,Czech:2019vih,Apolo:2020qjm}. 
	}:
	\begin{equation}
		\xi = 2\pi \sum_{n=0,\pm 1} \pqty{
				s_n L_n - \bar{s}_n \bar{L}_n
			},
	\end{equation}
	\begin{equation}
		s_{-1} = \frac{u_+ u_-}{u_+ - u_-},\quad
		s_0 = - \frac{u_+ + u_-}{u_+ - u_-},\quad
		s_{+1} = \frac{1}{u_+ - u_-}
	\end{equation}
	One need only exchange all $u\leftrightarrow v$ to obtain $\bar{s}_n$. The overall coefficient of $\xi$ (in this case, $2\pi$) is fixed by normalizing the temperature of the horizon, namely we demand that the surface gravity is $2\pi$ along the RT surface:
	\begin{equation}
		T_\xi = \frac{\kappa_\xi}{2\pi} = 1,
	\quad
		\kappa_\xi^2
		= (2\pi)^2
		= -\frac{1}{2}\,
			(\nabla_{\mu} \xi_{\nu})
			(\nabla^{\mu} \xi^{\nu})
	\end{equation}

@@

	We note that the $\pdd{z}$ term vanishes in the following two cases:
	\begin{itemize}[noitemsep]
	\item as $z\to 0$, i.e.~near the asymptotic boundary;
	\item within the whole plane: $
			x\sinh\theta
			= t\cosh\theta
		$ where the RT surface lies. 
	\end{itemize}

@@

@@

	Due to conformal symmetry, one can further drop the conformal factor $\Omega^2(\tau,\tilde{x})$ from the metric, while the thermal entropy of the system remains unchanged. We now have a flat annulus:
	\begin{equation}
		\tilde{\mcal{R}}
		= S^1_\beta \times [0,\tilde{L}],
	\quad
		\dd{\tilde{s}}^2
		= \frac{1}{\Omega^2} \dd{s}^2
		= \dd{\tau}^2 + \dd{\tilde{x}}^2
	\end{equation}
	This is the corresponding geometry of a thermal partition function, with a constant temperature $\beta_\tau = 2\pi$. 
	Now $\pdd{\tau}$ is truly a Killing vector of $\tilde{\mcal{R}}$. 
	In this case the thermal entropy is extensive: the total entropy $S_\mcal{A}$ is proportional to the size of the system:
	\begin{equation}
		S_\mcal{A}
%		= s\,\norm{f(\mcal{A})}_{\tilde{\mcal{R}}}
		= \frac{c}{6}\,\norm{f(\mcal{A})}_{\tilde{\mcal{R}}}
	\end{equation}
	Here $\norm{f(\mcal{A})}_{\tilde{\mcal{R}}}$ denotes the $\tilde{R}$ length of the interval $f(\mcal{A})$:
	\begin{equation}
		\norm{f(\mcal{A})}_{\tilde{\mcal{R}}}
		= \int_{f(\mcal{A})} \dd{\tilde{s}}
		= \int_{f(\mcal{A})} \frac{\dd{s}}{\Omega}
		= \int_{\mcal{A}} \frac{\dd{s}}{\Omega}
		= \int_{\mcal{A}} \frac{2\pi}{\beta(x)} \dd{x}
	\end{equation}
	We've used the fact the $f$ is a diffeomorphism, to reduce the integral back to an integral along $\mcal{A}$ in the original coordinates. 
	The entropy is thus given by:
	\begin{equation}
		S_\mcal{A}
		= \frac{c}{6}
			\int_{\mcal{A}} \frac{2\pi}{\beta(x)} \dd{x}
	\end{equation}

@@

\pagebreak
\section{Entanglement Entropy as Gravitational Charge}
	Ref: \textcite{Wald:1993nt,Iyer:1994ys,Iyer:1995kg}, \textcite{Lewkowycz:2013nqa}, and \textcite{Faulkner:2013ana}.

@@

	By definition the $\int_\gamma$ integration goes along the RT surface; we should be careful in the calculations and include the contributions from the $(\chi_\xi)_r \dd{r}$ component. 
	Alternatively, using Stokes' theorem, it can be done at the asymptotic infinity $r\to\infty$ instead; in this case we need only the $\phi$ component. These two integration contours give the same result:
	\begin{equation}
		\var{Q}_\xi
		= \frac{\var{T}}{2G} \pqty{
				L \coth LT - \frac{1}{T}
			}
	\end{equation}
	With a final integration w.r.t.~$\var{T}$ in the phase phase, we recover the resulting charge, which agrees with the entanglement entropy in \cite{Rangamani:2016dms} up to an overall constant shift:
	\begin{equation}
		Q_\xi = \frac{1}{2G} \log \frac{\sinh LT}{T}
	\end{equation}

@@

>>>>

	\item We then note that the periodicity can be restored by following the flow of $\xi$ starting from a codim-2 cutoff surface. This could be related to the \textit{restricted maximin} proposal of \cite{Marolf:2019bgj,Grado-White:2020wlb}. \sidenote{[With this cutoff the causal wedge again coincides with the entanglement wedge]}

@@

\section{Zero temperature: modular flow in the Poincar\'e patch}

\subsection{Review of the general recipe}
	A boundary interval $\mcal{A}$ defines a denisty matrix:
	\begin{equation}
		\rho_\mcal{A}
		= \tr_{\bar{\mcal{A}}} \rho
		= e^{-\mcal{H}_\mcal{A}}
	\end{equation}
	where $\bar{\mcal{A}}$ is the complement of $\mcal{A}$ and $\mcal{H}_\mcal{A}$ is the corresponding modular Hamiltonian. $\mcal{H}_\mcal{A}$ is generically nonlocal, but in a highly symmetric background (e.g.~AdS vacuum) it might be geometrically realized and corresponds to a symmetry of the underlying quantum field theory. This is the modular flow $\xi$.

	In such case there is a generalized Rindler transformation $f$ that maps the casual domain $\mcal{D}$ to a patch $\{(\tau,\tilde{x})\}$ in some noncompact generalized Rindler spacetime, with $\pdd{\tau}$ an isometry of the spacetime. Then we have:
	\begin{equation}
		\xi = 2\pi \pdd{\tau}
		= 2\pi \sum_i a^i H_i,
	\quad
		\tau \sim \tau + 2\pi i
		= \tau + i\beta_\tau,
	\quad
		\beta_\tau = 2\pi
	\end{equation}
	Here $H_i$ are the symmetry generators. The causal domain of dependence $\mcal{D}$ is mapped to some thermal state with normalized inverse temperature $\beta_\tau = 2\pi$. Here we choose to absorb such $2\pi$ into the definition of $\xi$, so the Euclidean periodicity w.r.t.~$\xi$ is simply $\beta_\xi = 1$, which means the surface gravity along the horizon is given by:
	\begin{equation}
		\kappa_\xi = \frac{2\pi}{\beta_\xi} = 2\pi,
	\quad
		\beta_\xi = 1
	\end{equation}

@@

\subsubsection{Matching bulk and boundary symmetries}
	In Poincar\'e $\mrm{AdS}_3/\mrm{CFT}_2$, the $H_i$'s are given by $L_n,\tilde{L}_n$ with $n=0,\pm 1$. At $z\to 0$ we recover the usual global $\mfrak{sl}(2,\mbb{R})$ generators:
	\begin{equation}
		z\to 0,
	\quad
		      L_n \to -u^{n+1} \pdd{u},\quad
		\tilde{L}_n \to -v^{n+1} \pdd{v},\quad
	n = 0,\pm 1
	\end{equation}

	To extend it to the bulk, think of the corresponding bulk isometries: translations, rotations, dilations and special conformal transformations:
	\begin{equation}
	\begin{aligned}
		P_\mu
		&= \pdd{\mu},\\
		m_{uv}
		&= x_\mu \pdd{\nu} - x_\nu \pdd{\mu}
		= u\pdd{u} - v\pdd{v},\\
		\Delta
		&= X^I \pdd{I}
		= u\pdd{u} + v\pdd{v} + z\pdd{z},\\
		K_\mu
		&= (z^2 + uv)\,\pdd{\mu}
			- 2\eta_{\mu\nu} x^\nu \Delta,
	\end{aligned}
	\end{equation}
	\begin{equation}
		K_u = (z^2 + uv)\,\pdd{u} - v \Delta,
	\quad
		K_v = (z^2 + uv)\,\pdd{v} - u \Delta
	\end{equation}
	By requiring that a linear combination of these generators reduce to the boundary $L_n$'s, one can uniquely extend the $L_n,\tilde{L}_n$'s into the bulk. The result for $L_n$'s is listed below; for $\tilde{L}_n$ one need only exchange all $u\leftrightarrow v$.
	\begin{equation}
		L_{-1} = -\pdd{u},
	\quad
		L_0 = -u\,\pdd{u} - \frac{z}{2}\,\pdd{z},
	\quad
		L_1 = -u^2\pdd{u} + z^2 \pdd{v}
			- uz\,\pdd{z}
	\end{equation}

@@

	The RT surface (spacelike geodesic of Poincar\'e AdS with both ends attached to the boundary) is parametrized in terms of its end points $(t,x)_\pm$; it's the intersection of a AdS ``sphere'' (hyperboloid) and a plane through its center:
	\begin{equation}
	\begin{aligned}
	\text{``sphere'':} &&&\mathsmaller{
		z^2
		+ \pqty{x - \frac{x_+ + x_-}{2}}^{\!\!2}
		- \pqty{t - \frac{t_+ + t_-}{2}}^{\!\!2}
		= \pqty{\frac{L_\infty}{2}}^{\!\!2}
		= \pqty{\frac{x_+ - x_-}{2}}^{\!\!2}
		- \pqty{\frac{t_+ - t_-}{2}}^{\!\!2}
	},\\
	&&& \text{i.e.}\ \,
		L^2
		= \pqty{x_+ - x_-}^2
			- \pqty{t_+ - t_-}^2
		= (u_+ - u_-)(v_+ - v_-),
	\\[1.5ex]
	\text{plane:} &&&
		\frac{x - \frac{1}{2} (x_+ + x_-)}{x_+ - x_-}
		= \frac{t - \frac{1}{2} (t_+ + t_-)}{t_+ - t_-}
	\end{aligned}
	\end{equation}
	Alternatively, it can be more compactly rewritten with $x^\mu \sim (t,x)$ or $(u,v)$; mind the various $\pm$ signs in the expressions: 
	\begin{equation}
	\begin{aligned}
		0 &= z^2
			+ \eta_{\mu\nu} (x - x_+)^\mu (x - x_-)^\nu \\
		& = z^2
			+ (x - x_+)(x - x_-)
			- (t - t_+)(t - t_-) \\
		&= z^2 + \frac{
				(u - u_+)(v - v_-)
				+ (v - v_+)(u - u_-)
			}{2}
	\end{aligned}
	\end{equation}

	The bulk modular generator vanishes along the RT surface, i.e.~the RT surface is the fixed point of the modular flow; with this condition we can fix the coefficients of the modular generator, up to an overall coefficient\footnote{
		One can compare this with similar results in the literature such as \cite{Lashkari:2016idm,Czech:2019vih,Apolo:2020qjm}. 
	}:
	\begin{equation}
		\xi = 2\pi \sum_{n=0,\pm 1} \pqty{
				s_n L_n - \tilde{s}_n \tilde{L}_n
			},
	\end{equation}
	\begin{equation}
		s_{-1} = \frac{u_+ u_-}{u_+ - u_-},\quad
		s_0 = - \frac{u_+ + u_-}{u_+ - u_-},\quad
		s_{+1} = \frac{1}{u_+ - u_-}
	\end{equation}
	One need only exchange all $u\leftrightarrow v$ to obtain $\tilde{s}_n$. The overall coefficient of $\xi$ (in this case, $2\pi$) is fixed by normalizing the temperature of the horizon, namely we demand that the surface gravity is $2\pi$ along the RT surface:
	\begin{equation}
		\beta_\xi = \frac{2\pi}{\kappa_\xi} = 1,
	\quad
		\kappa_\xi^2
		= (2\pi)^2
		= -\frac{1}{2}\,
			(\nabla_{\mu} \xi_{\nu})
			(\nabla^{\mu} \xi^{\nu})
	\end{equation}

@@

	We note that the $\pdd{z}$ term vanishes in the following two cases:
	\begin{itemize}[noitemsep,topsep=0pt]
	\item as $z\to 0$, i.e.~near the asymptotic boundary;
	\item within the whole plane: $
			x\sinh\theta
			= t\cosh\theta
		$ where the RT surface lies. 
	\end{itemize}

@@

	\sidenote{One oddity of this prescription is that the codim-1 cutoff geometry depends on the choice of the interval $\mcal{A}$}. This seems to suggest that there is no canonical way to embed the entire time-evolved deformed field theory into the bulk. The best we can do is to embed the causal diamond $\mcal{D}(\mcal{A})$ (\sidenote{and maybe $\mcal{D}(\mcal{A})$}), given an interval $\mcal{A}$. This is the price we have to pay for SSA to hold.

	Maybe we should not expect that the entire time-evolved deformed theory can be embedded into the bulk. \sidenote{We could look at the proof of $T\bar{T}$ duality and check if this is really implied. It is quite possible that we only need to embed the boundary theory at a specific time slice, and the trace flow arguments will go through.}

@@

	Due to conformal symmetry [\sidenote{what about the TTbar case? Perhaps the reason it still works is again, that the conformal factor vanishes at the endpoints, due to $2\pi$ periodicity of $\xi$}], one can further drop the conformal factor $\Omega^2(\tau,\tilde{x})$ from the metric, while the thermal entropy of the system remains unchanged. We now have a flat annulus:
	\begin{equation}
		\tilde{\mcal{R}}
		= S^1_\beta \times [0,\tilde{L}],
	\quad
		\dd{\tilde{s}}^2
		= \frac{1}{\Omega^2} \dd{s}^2
		= \dd{\tau}^2 + \dd{\tilde{x}}^2
	\end{equation}
	This is the corresponding geometry of a thermal partition function, with a constant temperature $\beta_\tau = 2\pi$. 
	Now $\pdd{\tau}$ is truly a Killing vector of $\tilde{\mcal{R}}$. 
	In this case the thermal entropy is extensive: the total entropy $S_\mcal{A}$ is proportional to the size of the system:
	\begin{equation}
		S_\mcal{A}
%		= s\,\norm{f(\mcal{A})}_{\tilde{\mcal{R}}}
		= \frac{c}{6}\,\norm{f(\mcal{A})}_{\tilde{\mcal{R}}}
	\end{equation}
	Here $\norm{f(\mcal{A})}_{\tilde{\mcal{R}}}$ denotes the $\tilde{R}$ length of the interval $f(\mcal{A})$:
	\begin{equation}
		\norm{f(\mcal{A})}_{\tilde{\mcal{R}}}
		= \int_{f(\mcal{A})} \dd{\tilde{s}}
		= \int_{f(\mcal{A})} \frac{\dd{s}}{\Omega}
		= \int_{\mcal{A}} \frac{\dd{s}}{\Omega}
		= \int_{\mcal{A}} \frac{2\pi}{\beta(x)} \dd{x}
	\end{equation}
	We've used the fact the $f$ is a diffeomorphism, to reduce the integral back to an integral along $\mcal{A}$ in the original coordinates. 
	The entropy is thus given by:
	\begin{equation}
		S_\mcal{A}
		= \frac{c}{6}
			\int_{\mcal{A}} \frac{2\pi}{\beta(x)} \dd{x}
	\end{equation}

@@

\pagebreak
\section{Finite temperature: modular flow in BTZ background}
\subsection{Entanglement Entropy as Gravitational Charge}
	Ref: \textcite{Wald:1993nt,Iyer:1994ys,Iyer:1995kg}, \textcite{Lewkowycz:2013nqa}, and \textcite{Faulkner:2013ana}.

@@

	By definition the $\int_\gamma$ integration goes along the RT surface; we should be careful in the calculations and include the contributions from the $(\chi_\xi)_r \dd{r}$ component. 
	Alternatively, using Stokes' theorem, it can be done at the asymptotic infinity $r\to\infty$ instead; in this case we need only the $\phi$ component. These two integration contours give the same result:
	\begin{equation}
		\var{Q}_\xi
		= \frac{\var{T}}{2G} \pqty{
				L \coth LT - \frac{1}{T}
			}
		= \var{S_\mcal{A}}
		= \var{\ave{\mcal{H}_\mcal{A}}}
	\end{equation}
	This is the holographic realization of the first law of entanglement.

	With a final integration w.r.t.~$\var{T}$ in the phase phase, we recover the resulting charge, which agrees with the entanglement entropy in \cite{Rangamani:2016dms} up to an overall constant shift:
	\begin{equation}
		Q_\xi = \frac{1}{2G} \log \frac{\sinh LT}{T}
	\end{equation}

@@

\subsection{Field theory side calculation}
